---
title: "Univariate state-space models"
subtitle: "PSAW II – Time Series Analysis Training Session"
author: "Mark Scheuerell"
date: "11 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(123)
```

## Topics

Univariate state-space models

1. state (process) model

2. observation model


## State-space models

Consist of 2 parts


## Part 1: State model

Describes the __true state of nature__ over time

```{r, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center'}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## boundaries
ss <- 5
nn <- 7
rr <- ss*3
cc <- ss*nn
## mid-points
xm <- ss/2 + seq(0,cc-ss,ss)
ymt <- rr - ss/2
ymb <- ss/2
## arrow locs
x0t <- seq(ss, by=2*ss, len=3)
x1t <- x0t + ss
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row: state
symbols(x=xm[c(1,3,5,7)], y=rep(ymt,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#488fdf",
        inches=FALSE, add=TRUE, lwd=3)
text("Truth", x=-ss, y=ymt, adj=c(0,0.5), xpd=NA,
     cex=2, col="#488fdf")
arrows(x0=x0t,x1=x1t,y0=ymt, col="#488fdf", lwd=3, length=0.12)
## Time or space
arrows(x0=ss/2, x1=cc-ss/2, y0=-ss/3+ss*2,
       length=0.12, lwd=3, xpd=NA)
text("Time or space", x=cc/2, y=-ss/2+ss*2, xpd=NA, pos=1, cex=2)
```


## Part 1: State model

States of nature might be:

 * animal's location
 
 * density of organisms
 
 * reproductive state
 

## Changes in the state of nature

We can think of changes in the state of nature being driven by a combination of

 * intrinsic (eg, fecundity), and

 * extrinsic factors (eg, temperature)


## Process errors

Some of the extrinsic drivers may be unknown

In time series modeling, we often call these unknown extrinsic factors _process errors_


## Part 2: Observation model

__Data = Truth + Error__
 
```{r obs_diag, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center'}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## arrow locs
x0t <- seq(ss, by=2*ss, len=3)
x1t <- x0t + ss
y0b <- rr - ss
y1b <- ss
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row: state
symbols(x=xm[c(1,3,5,7)], y=rep(ymt,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#488fdf",
        inches=FALSE, add=TRUE, lwd=3)
text("Truth", x=-ss, y=ymt, adj=c(0,0.5), xpd=NA,
     cex=2, col="#488fdf")
## arrows
arrows(x0=x0t,x1=x1t,y0=ymt, col="#488fdf", lwd=3, length=0.12)
## bottom row: obs
symbols(x=xm[c(1,3,5,7)], y=rep(ss/2,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#844870",
        inches=FALSE, add=TRUE, lwd=3)
text("Data", x=-ss, y=ss/2, adj=c(0,0.5), xpd=NA,
     cex=2, col="#844870")
## arrows
arrows(x0=xm[c(1,3,5,7)], y0=y0b, y1=y1b,
       col="#c10101", lwd=3, length=0.12)
## Time or space
arrows(x0=ss/2, x1=cc-ss/2, y0=-ss/3,
       length=0.12, lwd=3, xpd=NA)
text("Time or space", x=cc/2, y=-ss/2, xpd=NA, pos=1, cex=2)
```


## Observation (sampling) errors

Consider for a moment the very act of collecting data


