---
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
```


# Part 2: Observation model

## .purple-text[Stuff we observe] = .blue-text[Stuff nature does] &#177; .red-text[Stuff we miss]

---

# State model for species interactions

### Number today is a function of the .gray-text[number yesterday]

```{r MAR_diag_1, dpi=300, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center', echo=FALSE, warning=FALSE}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## boundaries
ss <- 5
nn <- 8
rr <- ss*3
cc <- ss*nn
## mid-points
xm <- seq(5,cc-ss,rr)
ymt <- rr - ss/2
ymb <- ss/2
## arrow locs
y0 <- rr - ss
y1 <- ss
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row
symbols(x=xm[2], y=ymt, rectangles=matrix(c(2*ss,ss),1,2),
        lty="solid",  bg="#656565", fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text("Number", x=xm[2], y=ymt, cex=1.8, col="#ffffff", pos=3)
text("yesterday", x=xm[2], y=ymt, cex=1.8, col="#ffffff", pos=1)
## arrows
arrows(x0=xm[2], y0=y0, y1=y1,
       col="#656565", lwd=3, length=0.12)
## bottom row: obs
symbols(x=xm[2], y=ymb, rectangles=matrix(c(2*ss,ss),1,2),
        lty="solid",  bg="#488fdf", fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text("Number", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=3)
text("today", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=1)
```

---

# State model for species interactions

### ...and the number of .red-text[predators], .orange-text[prey], .purple-text[competitors]

```{r MAR_diag_2, dpi=300, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center', echo=FALSE, warning=FALSE}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row
symbols(x=xm, y=rep(ymt,3),
        rectangles=matrix(c(2*ss,ss),3,2,byrow=TRUE),
        lty="solid",  bg=c("#c10101","#ff8100","#844870"),
        fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text(c("Predators","Prey","Competitors"),
     x=xm, y=rep(ymt,3), cex=1.7, col="#ffffff", pos=3)
text("yesterday", x=xm, y=rep(ymt,3), cex=1.7, col="#ffffff", pos=1)
## arrows
arrows(x0=ss*seq(2,6,2), x1=ss*seq(3,5),
       y0=y0, y1=y1,
       col=c("#c10101","#ff8100","#844870"), lwd=3, length=0.12)
## bottom row: obs
symbols(x=xm[2], y=ymb, rectangles=matrix(c(2*ss,ss),1,2),
        lty="solid",  bg="#488fdf", fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text("Number", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=3)
text("today", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=1)
```

---

# State model for species interactions

### ...and .green-text[external forces] at various times

```{r MAR_diag_3, dpi=300, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center', echo=FALSE, warning=FALSE}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row
symbols(x=xm[2], y=ymt, rectangles=matrix(c(2*ss,ss),1,2),
        lty="solid",  bg="#339933", fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text("External", x=xm[2], y=ymt, cex=1.8, col="#ffffff", pos=3)
text("forces", x=xm[2], y=ymt, cex=1.8, col="#ffffff", pos=1)
## arrows
arrows(x0=xm[2], y0=y0, y1=y1,
       col="#339933", lwd=3, length=0.12)
## bottom row: obs
symbols(x=xm[2], y=ymb, rectangles=matrix(c(2*ss,ss),1,2),
        lty="solid",  bg="#488fdf", fg=NA,
        inches=FALSE, add=TRUE, lwd=3)
text("Number", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=3)
text("today", x=xm[2], y=ymb, cex=1.8, col="#ffffff", pos=1)
```

---

# Applications of hierarchical models

.blue-text[
## Individual movements

## Species interactions

## Joint species distributions

## Meta-analyses
]
