---
title: "White noise & random walks"
subtitle: "PSAW II – Time Series Analysis Training Session"
author: "Mark Scheuerell"
date: "11 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(123)
```

## Topics

White noise

Random walks

Biased random walks


## White noise (WN)

A time series $\{w_t\}$ is discrete white noise if its values are

1. independent  

2. identically distributed with a mean of zero


## White noise (WN)

A time series $\{w_t\}$ is discrete white noise if its values are

1. independent  

2. identically distributed with a mean of zero

Note that distributional form for $\{w_t\}$ is flexible


## White noise (WN)

```{r, fig.cap="$w_t = 2e_t - 1; e_t \\sim \\text{Bernoulli}(0.5)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- rbinom(100, 1, 0.5) * 2 - 1
plot.ts(tt, ylab = expression(italic(w[t])))
acf(tt)
```


## Gaussian white noise

We often assume so-called _Gaussian white noise_, whereby

$$
w_t \sim \text{N}(0,\sigma^2)
$$


## Gaussian white noise

We often assume so-called _Gaussian white noise_, whereby

$$
w_t \sim \text{N}(0,\sigma^2)
$$

and the following apply as well

&nbsp; &nbsp; autocovariance:&nbsp; $\gamma_k =
    \begin{cases}
      \sigma^2 & \text{if } k = 0 \\
      0 & \text{if } k \geq 1
    \end{cases}$

&nbsp; &nbsp; autocorrelation: &nbsp; $\rho_k =
    \begin{cases}
      1 & \text{if } k = 0 \\
      0 & \text{if } k \geq 1
    \end{cases}$


## Gaussian white noise

```{r ex_gaussian_wn, fig.cap="$w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- rnorm(100)
plot.ts(tt, ylab = expression(italic(w[t])))
acf(tt)
```


## Random walk (RW)

A time series $\{x_t\}$ is a random walk if

1. $x_t = x_{t-1} + w_t$  

2. $w_t$ is white noise


## Random walk (RW)

```{r ex_rw, fig.cap="$x_t = x_{t-1} + w_t; w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- cumsum(rnorm(100))
plot.ts(tt, ylab = expression(italic(x[t])))
acf(tt)
```


## Biased random walk

A _biased random walk_ (or _random walk with drift_) is written as

$$
x_t = x_{t-1} + u + w_t
$$  

where $u$ is the bias (drift) per time step and $w_t$ is white noise


## Biased random walk

```{r ex_biased_rw, fig.cap="$x_t = x_{t-1} + 1 + w_t; w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
xx <- ww <- rnorm(100)
uu <- 1
for(t in 2:100) {
  xx[t] <- xx[t-1] + uu + ww[t]
}
plot.ts(xx, ylab = expression(italic(x[t])))
acf(tt)
```
