---
title: "Dynamic Factor Analysis"
subtitle: "PSAW II – Time Series Analysis Training Session"
author: "Mark Scheuerell"
date: "11 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
set.seed(123)
```

## Topics

Dynamic Factor Analysis (DFA)

* Forms of covariance matrix  

* Constraints for model fitting

* Interpretation of results


## Simple model for 2+ time series | Random walk observed with error

$$
y_{i,t} = x_{i,t} + a_i + v_{i,t} \\
x_{i,t} = x_{i,t-1} + w_{i,t}
$$

with

$v_{i,t} \sim \text{N}(0, R)$

$w_{i,t} \sim \text{N}(0, Q)$


## Random walk observed with error

$$
y_{1,t} = x_{1,t} + a_1 + v_{1,t} \\
y_{2,t} = x_{2,t} + a_2 + v_{2,t} \\
\vdots \\
y_{n,t} = x_{n,t} + a_2 + v_{n,t} \\
$$

$$
x_{1,t} = x_{1,t-1} + w_{1,t} \\
x_{2,t} = x_{2,t-1} + w_{2,t} \\
\vdots \\
x_{n,t} = x_{n,t-1} + w_{n,t}
$$


## Random walk observed with error | In matrix form

$$
\mathbf{y}_t = \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

with 

$\mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})$

$\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})$


##

```{r plot_many_ts, echo=FALSE, fig.align='center'}
NN <- 25
TT <- 30
MM <- 3
 
## MM x TT matrix of innovations
ww <- matrix(rnorm(MM*TT, 0, 1), MM, TT)
ww[,1] <- rnorm(MM, 0, sqrt(5))
## MM x TT matrix of scaled latent trends
xx <- t(scale(apply(ww,1,cumsum)))

## loadings matrix
ZZ <- matrix(runif(NN*MM, -1, 1), NN, MM)
diag(ZZ) <- rev(sort(abs(diag(ZZ))))
ZZ[upper.tri(ZZ)] <- 0
ZZ <- round(ZZ, 2)

## obs var
obs_var <- 0.2^2
## obs errors
ee <- t(MASS::mvrnorm(TT, matrix(0,NN,1), diag(obs_var,NN,NN)))
## NN x TT matrix of observed data
yy <- ZZ %*% xx + ee

clr <- viridis::plasma(NN, alpha=0.7, end=0.8)

vv <- sample(seq(NN), NN)

par(mfrow=c(5,5), mai=c(0.1,0.1,0,0), omi=c(0,0,0,0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd=2,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}
```


## Environmental time series

We often observe covariance among environmental time series, especially for those close to one another


##

```{r, echo=FALSE, fig.align='center'}
par(mfrow=c(5,5), mai=c(0.1,0.1,0,0), omi=c(0,0,0,0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd=2,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}

```

Are there some common patterns here?


## Common patterns in time series

```{r plot_dfa_trends, fig.align="center"}
## plot the trends
par(mfrow=c(1,3), mai=c(1.2,0,0,0), omi=rep(0.1,4))
clr <- viridis::plasma(MM, end=0.8)
for(i in 1:3) {
	plot.ts(xx[i,], lwd=3,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}
```


## State-space model | Ex: population structure

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$


## State-space model | Ex: Harbor seal population structure

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 \\
 y_4 \\
 y_5 
\end{bmatrix}_t =
\begin{bmatrix}
 1 & 0 & 0 \\
 0 & 1 & 0 \\
 0 & 1 & 0 \\
 0 & 0 & 1 \\
 0 & 0 & 1 \\
\end{bmatrix} \times
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t +
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5 
\end{bmatrix} +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 \\
 v_4 \\
 v_5 
\end{bmatrix}_t
$$

$$
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t =
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_{t-1} +
\begin{bmatrix}
 w_{JF} \\
 w_N \\
 w_S
\end{bmatrix}_t
$$


## Finding common patterns

What if our observations were instead a mixture of 2+ states?

For example, we sampled haul-outs located between several breeding sites


## Mixtures of states

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 \\
 y_4 \\
 y_5 
\end{bmatrix}_t =
\begin{bmatrix}
 0.8 & 0.2 & 0 \\
 0.2 & 0.7 & 0.1 \\
 0 & 0.9 & 0.1 \\
 0 & 0.3 & 0.7 \\
 0 & 0.1 & 0.9 \\
\end{bmatrix} \times
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t +
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5 
\end{bmatrix} +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 \\
 v_4 \\
 v_5 
\end{bmatrix}_t
$$

$$
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t =
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_{t-1} +
\begin{bmatrix}
 w_{JF} \\
 w_N \\
 w_S
\end{bmatrix}_t
$$


## Finding common patterns

What if our observations were a mixture of states, but we didn't know how many or the weightings?

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What are the dimensions of $\mathbf{Z}$?

What are the elements within $\mathbf{Z}$?


## Dynamic Factor Analysis (DFA)

DFA is a _dimension reduction_ technique, which models $n$ observed time series as a function of $m$ hidden states (patterns), where $n \gg m$


## Dynamic Factor Analysis (DFA) | State-space form

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

data: $\mathbf{y}_t$ is $n \times 1$

loadings: $\mathbf{Z}$ is $n \times m$ with $n > m$

states: $\mathbf{x}_t$ is $m \times 1$


## Dimension reduction | Principal Components Analysis (PCA)

Goal is to reduce some large number of correlated variates into a few uncorrelated factors


## Principal Components Analysis (PCA)

Calculating the principal components requires us to estimate the covariance of the data

$$
\text{PC} = \text{eigenvectors}(\text{cov}(\mathbf{y}))
$$

There will be $n$ principal components (eigenvectors) for an $n \times T$ matrix $\mathbf{y}$

We reduce the dimension by selecting a subset of the components that explain much of the variance (eg, the first 2)


## Principal Components Analysis (PCA)

```{r}
nn <- 100
xx <- rnorm(nn,10,1)
yy <- xx + rnorm(nn, 5)
xx <- xx - mean(xx)
yy <- yy - mean(yy)

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))

plot(xx, yy, pch = 16, las = 1,
     ylab ="Y", xlab = "X")

text(-2.5, 3,
     substitute(rho==x, list(x = round(cor(cbind(xx,yy))[2,1], 2))))
```


## Principal Components Analysis (PCA)

```{r}
pca <- prcomp(cbind(xx,yy))

pc1 <- pca$rotation[,1]
pc2 <- matrix(c(0,-1,1,0), 2, 2) %*% matrix(pc1, 2, 1)

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))

plot(xx, yy, pch = 16, las = 1,
     ylab ="Y", xlab = "X")
arrows(0, 0, pc1[1], pc1[2],
       col = "blue", lwd = 3, length = 0.1)
text(pc1[1], pc1[2], "PC1",
     pos = 3, col = "blue")
arrows(0, 0, pc2[1], pc2[2],
       col = "darkred", lwd = 3, length = 0.1)
text(pc2[1], pc2[2], "PC2",
     pos = 1, col = "darkred")
```


## Principal Components Analysis (PCA)

```{r}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))

plot(pca$x[,1], pca$x[,2], pch = 16, las = 1,
     ylab ="PC2", xlab = "PC1")

text(-3, 1.2,
     substitute(rho==x, list(x = round(cor(pca$x)[2,1], 2))))
```


## Relationship between PCA & DFA

We need to estimate the covariance in the data $\mathbf{y}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t, ~ \text{with} ~ \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})
$$

so

$$
\text{cov}(\mathbf{y}_t) = \mathbf{Z} \text{cov}(\mathbf{x}_t) \mathbf{Z}^\top + \mathbf{R}
$$
In PCA, we require $\mathbf{R}$ to be diagonal, but not so in DFA


## Principal Components Analysis (PCA) | Forms for $\mathbf{R}$ with $n = 4$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & 0 & 0 & 0 \\
 0 & \sigma & 0 & 0 \\
 0 & 0 & \sigma & 0 \\
 0 & 0 & 0 & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & 0 \\
 0 & 0 & \sigma_3 & 0 \\
 0 & 0 & 0 & \sigma_4
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA) | Forms for $\mathbf{R}$ with $n = 4$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & 0 & 0 & 0 \\
 0 & \sigma & 0 & 0 \\
 0 & 0 & \sigma & 0 \\
 0 & 0 & 0 & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & 0 \\
 0 & 0 & \sigma_3 & 0 \\
 0 & 0 & 0 & \sigma_4
\end{bmatrix}
$$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & \gamma & \gamma & \gamma \\
 \gamma & \sigma & \gamma & \gamma \\
 \gamma & \gamma & \sigma & \gamma \\
 \gamma & \gamma & \gamma & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & \gamma_{2,4} \\
 0 & 0 & \sigma_3 & 0 \\
 0 & \gamma_{2,4} & 0 & \sigma_4
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA)

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What form should we use for $\mathbf{Z}$?

$$
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 \\
 z_4 \\
 z_5
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3} \\
 z_{1,4} & z_{2,4} \\
 z_{1,5} & z_{2,5}
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} & z_{3,1} \\
 z_{1,2} & z_{2,2} & z_{3,2} \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 z_{1,4} & z_{2,4} & z_{3,4} \\
 z_{1,5} & z_{2,5} & z_{3,5}
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA)

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What form should we use for $\mathbf{Z}$?

$$
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 \\
 \vdots \\
 z_5
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3} \\
 \vdots & \vdots \\
 z_{1,n} & z_{2,n}
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} & z_{3,1} \\
 z_{1,2} & z_{2,2} & z_{3,2} \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 \vdots & \vdots & \vdots \\
 z_{1,n} & z_{2,n} & z_{3,n}
\end{bmatrix}
$$

We'll use model selection criteria to choose (eg, AICc)


## Fitting DFA models

It turns out that there are an infinite number of combinations of $\mathbf{Z}$ and $\mathbf{x}$ that will equal $\mathbf{y}$

Therefore we need to impose some constraints on the model


## Constraints on DFA models | The offset $\mathbf{a}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 \vdots \\
 a_n
\end{bmatrix}
$$

We will set the first $m$ elements of $\mathbf{a}$ to 0


## Constraints on DFA models | The offset $\mathbf{a}$

For example, if $n = 5$ and $m = 2$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
$$


## Constraints on DFA models | The offset $\mathbf{a}$

For example, if $n = 5$ and $m = 2$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 0 \\
 0 \\
 0
\end{bmatrix}
$$

Note, however, that this causes problems for the EM algorithm so we will often de-mean the data and set $a_i = 0$ for all $i$


## Constraints on DFA models | The loadings $\mathbf{Z}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & z_{2,1} & \dots & z_{m,1} \\
 z_{1,2} & z_{2,2} & \dots & z_{m,2} \\
 z_{1,3} & z_{2,3} & \dots & z_{m,3} \\
 \vdots & \vdots & \ddots & z_{m,4} \\
 z_{1,n} & z_{2,n} & \dots & z_{m,n}
\end{bmatrix}
$$

We will set the upper right triangle of $\mathbf{Z}$ to 0


## Constraints on DFA models | The loadings $\mathbf{Z}$

For example, if $n = 5$ and $m = 3$

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & 0 & 0 \\
 z_{1,2} & z_{2,2} & 0 \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 z_{1,4} & z_{2,3} & z_{3,4} \\
 z_{1,5} & z_{2,5} & z_{3,5}
\end{bmatrix}
$$

For the first $m - 1$ rows of $\mathbf{Z}$, $z_{i,j} = 0$ if $j > i$


## Constraints on DFA models | The loadings $\mathbf{Z}$

An additional constraint is necessary in a Bayesian context

$$
\mathbf{Z} =
\begin{bmatrix}
 \underline{z_{1,1}} & 0 & 0 \\
 z_{1,2} & \underline{z_{2,2}} & 0 \\
 z_{1,3} & z_{2,3} & \underline{z_{3,3}} \\
 z_{1,4} & z_{2,3} & z_{3,4} \\
 z_{1,5} & z_{2,5} & z_{3,5}
\end{bmatrix}
$$

Diagonal of $\mathbf{Z}$ is positive: $z_{i,j} > 0$ if $i = j$


## Constraints on DFA models | The state variance $\mathbf{Q}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$

We will set $\mathbf{Q}$ equal to the Identity matrix $\mathbf{I}$


## Constraints on DFA models | The state variance $\mathbf{Q}$

For example, if $m = 4$

$$
\mathbf{Q} =
\begin{bmatrix}
 1 & 0 & 0 & 0 \\
 0 & 1 & 0 & 0 \\
 0 & 0 & 1 & 0 \\
 0 & 0 & 0 & 1
\end{bmatrix}
$$

This allows our random walks to have a _lot_ of flexibility


## Dynamic Factor Analysis (DFA) | Including $p$ covariates

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \underline{\mathbf{D} \mathbf{d}_t} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$\mathbf{d}_t$ is a $p \times 1$ vector of covariates at time $t$

$\mathbf{D}$ is an $n \times p$ matrix of covariate effects


## Dynamic Factor Analysis (DFA) | Form for $\mathbf{D}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \underline{\mathbf{D}} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

Careful thought must be given _a priori_ as to the form for $\mathbf{D}$

Should the effect(s) vary by site, species, etc?


## Dynamic Factor Analysis (DFA) | Form for $\mathbf{D}$

For example, given 2 covariates, $\text{Temp}$ and $\text{Salinity}$

$$
\mathbf{D} =
\underbrace{
\begin{bmatrix}
 d_{\text{Temp}} & d_{\text{Salinity}} \\
 d_{\text{Temp}} & d_{\text{Salinity}} \\
 \vdots & \vdots \\
 d_{\text{Temp}} & d_{\text{Salinity}} \\
\end{bmatrix}
}_{\text{effects same by site/species}}
~ \text{or} ~
\mathbf{D} =
\underbrace{
\begin{bmatrix}
 d_{\text{Temp}, 1} & d_{\text{Salinity}, 1} \\
 d_{\text{Temp}, 2} & d_{\text{Salinity}, 2} \\
 \vdots & \vdots \\
 d_{\text{Temp}, n} & d_{\text{Salinity}, n} \\
\end{bmatrix}
}_{\text{effects differ by site/species}}
$$


## A note on model selection

Earlier we saw that we could use model selection criteria to help us choose among the different forms for $\mathbf{Z}$

However, caution must be given when comparing models with and without covariates, and varying numbers of states


## A note on model selection

Think about the model form

$$
\mathbf{y}_t = \mathbf{Z} \underline{\mathbf{x}_t} + \mathbf{a} + \mathbf{D} \underline{\mathbf{d}_t} + \mathbf{v}_t \\
$$

$\mathbf{x}_t$ is an _undetermined_ random walk

$\mathbf{d}_t$ is a _predetermined_ covariate

Unless $\mathbf{d}$ is highly correlated with $\mathbf{y}$, then the inclusion of a state $\mathbf{x}$ will be favored over $\mathbf{d}$


## A note on model selection

Thus, work out fixed effects (covariates) while keeping the random effects (states) constant, and vice versa

For example, compare data support for models with different combinations of covariates, only one state ($m$ = 1), and a "diagonal and equal" $\mathbf{R}$


## Interpreting DFA results

Recall that we had to constrain the form of $\mathbf{Z}$ to fit the model

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & 0 & \dots & 0 \\
 z_{1,2} & z_{2,2} & \ddots & 0 \\
 \vdots & \vdots & \ddots & 0 \\
 \vdots & \vdots & \vdots & z_{m,m}  \\
 \vdots & \vdots & \vdots & \vdots \\
 z_{1,n} & z_{2,n} & z_{3,n} & z_{m,n} 
\end{bmatrix}
$$

So, the 1st common factor is determined by the 1st variate, the 2nd common factor by the first two variates, etc.


## Interpreting DFA results

To help with this, we can use a _basis rotation_ to maximize the loadings on a few factors

If $\mathbf{H}$ is an $m \times m$ non-singular matrix, these 2 DFA models are equivalent

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{H}^{-1} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{H} \mathbf{x}_t = \mathbf{H} \mathbf{x}_{t-1} + \mathbf{H} \mathbf{w}_t
$$

How should we choose $\mathbf{H}$?


## Basis rotation | Varimax

A _varimax_ rotation will maximize the variance of the loadings in $\mathbf{Z}$ along a few of the factors


## PCA of 5 wines with 8 attributes

```{r}
dat <- read.csv("wine_data_for_PCA.csv")
colnames(dat)[2:3] <- c("For meat", "For dessert")

pca <- prcomp(dat, scale. = TRUE)

loads <- pca$rotation
if(loads[1,2] < 0) { loads[,2] <- -1 * loads[,2] }

load1 <- loads[,1]
load2 <- loads[,2]

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 1.5, 0, 1.5))

poz <- c(3, 1, 3, 3, 3, 4, 2)
  
plot(load1, load2, pch = 16,
     xlim = c(min(load1), max(load1)),
     ylim = c(min(load2), max(load2)),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(h = 0, col = "darkgray")
abline(v = 0, col = "darkgray")
for(i in 1:ncol(dat)) {
  text(load1[i], load2[i], colnames(dat)[i],
       pos = poz[i], xpd = NA)
}
text(0, min(load2) * 1.05, "X2", pos = 1, xpd = NA, col = "darkgray")
text(min(load1) * 1.05, 0, "X1", pos = 2, xpd = NA, col = "darkgray")
```


## Rotated loadings

```{r}
theta <- 165 * pi / 180
basis_rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
  
new_x <- t(basis_rot %*% matrix(c(1,0,-1,0),2,2))
new_y <- t(basis_rot %*% matrix(c(0,1,0,-1),2,2))

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 1.5, 0, 1.5))

plot(load1, load2, pch = 16,
     xlim = c(min(load1), max(load1)),
     ylim = c(min(load2), max(load2)),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(h = 0, col = "darkgray")
abline(v = 0, col = "darkgray")
## new axes
lines(new_x[,1], new_x[,2], col = "blue")
lines(new_y[,1], new_y[,2], col = "blue")
## labels
for(i in 1:ncol(dat)) {
  text(load1[i], load2[i], colnames(dat)[i],
       pos = poz[i], xpd = NA)
}
text(0, min(load2) * 1.05, "X2", pos = 1, xpd = NA, col = "darkgray")
text(min(load1) * 1.05, 0, "X1", pos = 2, xpd = NA, col = "darkgray")

theta <- 345 * pi / 180
basis_rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
text((basis_rot %*% matrix(c(0,par()$usr[3]),2,1))[1], min(load2) * 1.05,
     "R2", pos = 1, xpd = NA, col = "blue")
text(min(load1) * 1.05, (basis_rot %*% matrix(c(par()$usr[1],0),2,1))[2],
     "R1", pos = 2, xpd = NA, col = "blue")

```


## Fitting DFA models

Estimating the parameters in a DFA model can be challenging and time intensive

Methods include

 * maximum likelihood (eg, [Chap 10 in MARSS Manual](https://cran.r-project.org/web/packages/MARSS/vignettes/UserGuide.pdf))
 
 * Bayesian (eg, [Stan](https://cran.r-project.org/web/packages/bayesdfa/index.html) or [greta](https://mdscheuerell.github.io/gretaDFA/))
 