---
title: "Important time series operators"
subtitle: "PSAW II – Time Series Analysis Training Session"
author: "Mark Scheuerell"
date: "11 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
set.seed(123)
```


## Important operators

This slide deck defines these 2 important operators:

* _backshift shift operator_ ($\mathbf{B}$)

* _difference operator_ ($\nabla$) 


## The backshift shift operator

The _backshift shift operator_ ($\mathbf{B}$) is an important function in time series analysis, which we define as

$$
\mathbf{B} x_t = x_{t-1}
$$

or more generally as

$$
\mathbf{B}^k x_t = x_{t-k}
$$

## The backshift shift operator

For example, a random walk with

$$
x_t = x_{t-1} + w_t
$$

can be written as

$$
\begin{align}
  x_t &= \mathbf{B} x_t + w_t \\
  x_t - \mathbf{B} x_t &= w_t \\
  (1 - \mathbf{B}) x_t &= w_t \\
  x_t &= (1 - \mathbf{B})^{-1} w_t
\end{align}
$$


## The difference operator

The _difference operator_ ($\nabla$) is another important function in time series analysis, which we define as

$$
\nabla x_t = x_t - x_{t-1}
$$


## The difference operator

The _difference operator_ ($\nabla$) is another important function in time series analysis, which we define as

$$
\nabla x_t = x_t - x_{t-1}
$$

For example, first-differencing a random walk yields white noise

$$
\begin{align}
  \nabla x_t &= x_{t-1} + w_t \\
  x_t - x_{t-1} &= x_{t-1} + w_t - x_{t-1}\\
  x_t - x_{t-1} &= w_t\\
\end{align}
$$


## Differencing a biased random walk

First-differencing a biased random walk yields a constant mean (level) $u$ plus white noise

$$
\begin{align}
  \nabla x_t &= x_{t-1} + u + w_t \\
  x_t - x_{t-1} &= x_{t-1} + u + w_t - x_{t-1} \\
  x_t - x_{t-1} &= u + w_t
\end{align}
$$


## Differencing a biased random walk

```{r ex_diff_biased_rw, fig.cap="$x_t = x_{t-1} + 1 + w_t; w_t \\sim \\text{N}(0,1)$"}
xx <- arima.sim(model=list(ar=0.5, sd=0.1), n=100)

par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(diff(xx), ylab = expression(nabla~italic(x[t])))
acf(diff(xx))
```


## The difference operator

The difference operator and the backshift operator are related

$$
\nabla^k = (1 - \mathbf{B})^k
$$


## The difference operator

The difference operator and the backshift operator are related

$$
\nabla^k = (1 - \mathbf{B})^k
$$

For example

$$
\begin{align}
  \nabla x_t &= (1 - \mathbf{B})x_t \\
  x_t - x_{t-1} &= x_t - \mathbf{B} x_t \\
  x_t - x_{t-1} &= x_t - x_{t-1}
\end{align}
$$


## Differencing to remove a trend

Differencing is a simple means for removing a trend

The 1st-difference removes a linear trend; a 2nd-difference would remove a quadratic trend, etc.


## Differencing to remove a trend

```{r diff_linear, fig.align="center"}
## create biased RW
rr <- ww <- rnorm(50)
for(t in 2:50) {
  rr[t] <- 0.3 + rr[t-1] + ww[t]
}

par(mfrow = c(2,1), mai = c(0.5,0.8,0.1,0), omi=c(0,0,0,0))

## raw data
plot.ts(rr, las = 1,
        ylab = expression(italic(x[t])))
## first difference
plot.ts(diff(rr), las = 1,
        ylab = expression(nabla~italic(x[t])))
```


## Differencing to remove seasonality

Differencing is a simple means for removing a seasonal effect

Using a 1st-difference with $k = period$ removes both trend & seasonal effects


## Differencing to remove seasonality

```{r diff_season, fig.align="center"}
library(MARSS)
data(lakeWAplankton)
lwa <- lakeWAplanktonTrans
par(mfrow = c(2,1), mai = c(0.5,0.8,0.1,0), omi=c(0,0,0,0))

## raw data
plot.ts(lwa[,"Temp"], las = 1,
        ylab = "Temperature")
## first difference
plot.ts(diff(lwa[,"Temp"], 12), las = 1,
        ylab = expression(nabla~Temperature))
```

